---
title: "Milestone Report - Week 2"
subtite: "Capstone Project"
author: "Jakob Schelbert"
date: "13. May 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

-------------------------------------------------------------------------------

The goal of this project is just to display that you've gotten used to working with the data and that you are on track to create your prediction algorithm. 
Please submit a report on R Pubs (http://rpubs.com/) that explains your exploratory analysis and your goals for the eventual app and algorithm. 

This document should be concise and explain only the major features of the data you have identified and briefly summarize your plans for creating the prediction algorithm and Shiny app in a way that would be understandable to a non-data scientist manager. 
You should make use of tables and plots to illustrate important summaries of the data set. 
The motivation for this project is to: 
1. Demonstrate that you've downloaded the data and have successfully loaded it in.
2. Create a basic report of summary statistics about the data sets.
3. Report any interesting findings that you amassed so far.
4. Get feedback on your plans for creating a prediction algorithm and Shiny app. 

Review criteria
    Does the link lead to an HTML page describing the exploratory analysis of the training data set?
    Has the data scientist done basic summaries of the three files? Word counts, line counts and basic data tables?
    Has the data scientist made basic plots, such as histograms to illustrate features of the data?
    Was the report written in a brief, concise style, in a way that a non-data scientist manager could appreciate?

-------------------------------------------------------------------------------




# Executive summary


# Introduction
Note that the execution of this R Markdown file requires a high amount of memory (16GB or more).


# Explorative analysis
```{r libraries}
library(tm)
library(ngram)
library(dplyr)
library(stringr)
library(ggplot2)
```


```{r load_data}
en_us <- VCorpus(DirSource("data/final/en_US/", encoding = "UTF-8"))
```






# Outline for prediction algorithm
In the next weeks we will implement a prediction algorithm that is able to suggest the next word to be typed based on the words already provided by the user.



# Way forward
In the next weeks we have quite some tasks before us that are described in the following chapter.

## Gathering additional data
While the originally provided data is a good starting point, more data will very likely enable us to build more accurate models.
In addition, there might be features in the current data that are not general and do not occur in other data sources.
To reduce the risk of building our models on biased data, we will search for more data which we can use in our analysis and development.

## Building the model
The main part of the project will be focused on the development of a model that makes accurate predictions.
In the academic literature there exist many approaches to tackle the problem.
Our goal is to start with very basic models and successively enhance our existing model to increase accuracy, response time and decrease model size.

## Fine tuning
A crucial and potentially tricky part of designing the algorithm will be the part of finding the right balance between model size, accuracy and response time for prediction.
With close alignment with our project sponsor we will run extensive studies of influencing parameters so at the end a model is found that suffices all requirements.

## Extension: Additional languages
While the main goal of the project will be to build a model based on the English language, the basic approach will also be feasible for other languages.
If time permits it, the next language we will incorporate is German (since its the mother tonge of the author).


# Summary
